---
title: "Case study- Acute Kidney injury comparison in Antiviral Treatment Regimens In Patients With Hepatitis C"
author: "Gilead Tutorial"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    theme: readable
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, eval=FALSE)
```

## NOTES

TO ADD

Code currently isn't runnable- editing in progress


## Introduction
Real-world evidence (RWE) studies in pharma often rely on traditional regression methods to analyze outcomes. However, causal inference requires more than a p-value or an adjusted hazard ratio (HR). The **causal inference roadmap** provides a structured framework to design analyses that yield **causally interpretable results**.

This tutorial will:
- Explain what an **estimand** is and why it matters in causal inference.
- Highlight **limitations of the Cox proportional hazards model and HR**.
- Introduce alternative estimands and estimation methods (e.g., **TMLE, G-computation**).
- Walk through a **case study** on Hepatitis C drug treatment and kidney injury.
- Provide **R code** to execute causal inference analyses using propensity score methods, TMLE, and diagnostics.

## The Causal Inference Roadmap & Estimand Selection
Modern causal inference recommends designing studies like a **target trial** by clearly specifying:
- **Causal question** (e.g., "Does Treatment X reduce 12-month kidney injury risk?")
- **Estimand** (e.g., **absolute risk difference**, survival probability, RMST)
- **Estimator** (e.g., **G-computation, TMLE, inverse probability weighting (IPW)**)

### Defining the Estimand
**Example:**
- "What is the **12-month risk difference** in kidney injury between treated vs. untreated patients?"
- **Estimand:** Absolute risk difference (Risk[treated] - Risk[untreated])
- **Estimators:** TMLE, G-computation, IPW

```{r}
# Example: Simulating hypothetical risk difference estimand
risk_treated <- 0.05  # 5% incidence
risk_untreated <- 0.08 # 8% incidence
risk_difference <- risk_treated - risk_untreated
risk_difference
```

## Issues with the Hazard Ratio (HR)
### Why HR May Be Misleading
- **Assumes Proportional Hazards**: HR is assumed constant over time.
- **Selection Bias**: HRs condition on survival, introducing bias.
- **Lack of Interpretability**: HR does not directly translate into **risk reduction**.

## Alternative Estimands & Methods
| **Estimand** | **Definition** | **Use Case** |
|-------------|--------------|--------------|
| **Risk Difference (RD)** | Difference in cumulative incidence between groups | Intuitive, direct impact on clinical decision-making |
| **Risk Ratio (RR)** | Ratio of cumulative incidence between groups | Epidemiologic comparisons |
| **Restricted Mean Survival Time (RMST)** | Difference in event-free time | Non-proportional hazards settings |

## Addressing Selection Bias & Informative Censoring
Selection bias occurs when loss to follow-up is related to **both treatment and outcome**.
### **Inverse Probability of Censoring Weighting (IPCW)**
```{r}
library(ipw)
censor_model <- glm(censor ~ age + diabetes + treatment, family = binomial, data = mydata)
mydata$weights <- 1 / predict(censor_model, type = "response")
```

### **TMLE for Informative Censoring**
```{r}
library(survtmle)
tmle_fit <- survtmle(ftime = mydata$time, ftype = mydata$event, trt = mydata$treat,
                     adjustVars = mydata[, c("age", "diabetes")],
                     t0 = 730,
                     glm.trt = "age + diabetes",
                     glm.ftime = "age + diabetes + treat",
                     glm.ctime = "age + diabetes")
summary(tmle_fit)
```

## Implementing Targeted Learning (TMLE) in R
### **Step 1: Specify Super Learner Library**
```{r}
library(SuperLearner)
SL.library <- c("SL.glm", "SL.glmnet", "SL.ranger", "SL.xgboost")
```

### **Step 2: TMLE Analysis**
```{r}
library(tmle)
tmle_fit <- tmle(Y = mydata$event, A = mydata$treatment, W = mydata[, c("age", "diabetes")],
                 family = "binomial", Q.SL.library = SL.library, g.SL.library = SL.library)
summary(tmle_fit)
```

## Diagnostics for TMLE
- **Super Learner Model Weights**
```{r}
print(tmle_fit$g$SL.weights)
print(tmle_fit$Q$SL.weights)
```

- **Comparison with Propensity Score Methods**
```{r}
library(MatchIt)
ps_model <- glm(treatment ~ age + diabetes, data = mydata, family = binomial)
ps <- predict(ps_model, type = "response")
match <- matchit(treatment ~ age + diabetes, data = mydata, method = "nearest", caliper = 0.2)
summary(match)
```

## Conclusion
- Always **define the estimand** before choosing an estimation method.
- Avoid **hazard ratios** as primary causal estimands.
- **Use absolute risk measures** (e.g., **risk difference, RMST**).
- **TMLE is robust** to model misspecification and provides better efficiency.

## References
- Hernán MA. *The hazards of hazard ratios.* Epidemiology. 2010.
- Snowden JM, Rose S, Mortimer KM. *G-computation for causal inference.* Am J Epidemiol. 2011.
- Gruber S, van der Laan MJ. *An R Package for Targeted Maximum Likelihood Estimation.*
