---
title: "HCV → AKI: 90‑Day Primary & ITT Sensitivity (Continuous‑Time TMLE via *concrete*)"
author: "Your Name Here"
output:
html_document:
toc: true
toc_depth: 2
---

> **Goal.** Re‑estimate the 90‑day cumulative risk of first acute kidney injury (AKI) for sofosbuvir‑containing vs non‑sofosbuvir DAA regimens using the **`concrete`** package’s one‑step *continuous‑time* TMLE:
>
> * **Primary estimand – Hypothetical *no‑switch***: censor subjects at first regimen switch (treated as right‑censoring) and target the counterfactual risk *if everyone remained on their index regimen*.
> * **Sensitivity estimand – Treatment‑policy (ITT)**: ignore switching (no extra censoring) to estimate the effect of *starting* a SOF‑containing vs non‑SOF regimen.
>
> `concrete` targets *cause‑specific absolute risks*; here we have a single cause (AKI) plus right‑censoring.
>
> The same simulated dataset (`sim_hcv_aki_complex.csv`) is analysed twice—once with switch → censoring (primary) and once with switch ignored (ITT).

```{r setup, message=FALSE, warning=FALSE}
# Core tidyverse + utilities
library(tidyverse)
library(here)

# concrete & friends
library(concrete)  # one‑step continuous‑time TMLE
library(SuperLearner)  # leveraged internally by concrete
```

\## 1   Load simulated data

```{r import}
# Load dataset (generated previously by DGP.R)
path <- here("data/sim_hcv_aki_complex.csv")
if (!file.exists(path)) stop("Data file not found: ", path)

df0 <- read_csv(path, show_col_types = FALSE)

# Glimpse key columns
print(names(df0))
```

**Assumed columns (modify if different)**

| Variable      | Meaning                                                           |
| ------------- | ----------------------------------------------------------------- |
| `A`           | Treatment indicator (1 = SOF, 0 = non‑SOF)                        |
| `follow_time` | min(actual event time, censoring time, switch +30 d)              |
| `event`       | 1 = AKI, 0 = no AKI by `follow_time`                              |
| `switch`      | 1 = regimen switch occurred at `follow_time`                      |
| covariates    | `age`, `sex_male`, `ckd`, `cirrhosis`, `diabetes`, `hypertension` |

We derive two analysis datasets:

* **No‑switch (primary)** – treat switching as right‑censoring.
* **ITT (sensitivity)** – ignore switching (switchers keep their observed follow‑up).

```{r derive-datasets}
max_follow <- 90  # analysis horizon (days)

make_concrete_input <- function(data, censor_at_switch = TRUE) {
  data %>%
    mutate(
      # event time is bounded by horizon
      time   = pmin(follow_time, max_follow),
      status = case_when(
        event == 1 & follow_time <= max_follow ~ 1,                # AKI
        censor_at_switch & switch == 1 & follow_time <= max_follow ~ 0,  # censored at switch
        TRUE ~ 0                                                   # administrative censor
      )
    ) %>%
    # Select columns needed by concrete (time, status, treatment + covariates)
    select(time, status, treatment, age, sex_male, ckd, cirrhosis, diabetes, hypertension) %>%
    # Drop any incomplete cases (concrete cannot handle missing)
    drop_na()
}

data_ns  <- make_concrete_input(df0, censor_at_switch = TRUE)   # primary

data_itt <- make_concrete_input(df0, censor_at_switch = FALSE)  # sensitivity
```

\## 2   Specify analysis arguments
`concrete` requires a *static* pair of interventions (`0:1`) and target times. We target the single horizon `t = 90` days.

```{r concrete-args}
# Common pieces
my_SL  <- list(V = 10)      # 10‑fold CV for SuperLearner
T0_vec <- c(max_follow)     # vector of target times (length 1)

# Function to build ConcreteArgs given a dataframe
make_args <- function(dat) {
  formatArguments(
    DataTable   = dat,
    EventTime   = "time",
    EventType   = "status",
    Treatment   = "trt",
    Intervention= 0:1,           # treat‑none vs treat‑all
    TargetTime  = T0_vec,
    TargetEvent = 1,             # only one cause of failure (AKI)
    CVArg       = my_SL,
    Verbose     = FALSE
  )
}

args_ns  <- make_args(data_ns)
args_itt <- make_args(data_itt)
```

\## 3   Run one‑step TMLE with `doConcrete()`

```{r run-concrete}
set.seed(202506)
fit_ns  <- doConcrete(args_ns)
fit_itt <- doConcrete(args_itt)
```

\## 4   Extract 90‑day risk differences & ratios

```{r extract-out}
# helper to pull RD & RR
get_RD_RR <- function(fit) {
  out <- getOutput(fit, Estimand = c("RD", "RR"), Simultaneous = FALSE)
  out %>%
    rename(Estimate = `Pt Est`, SE = se) %>%
    mutate(`95% CI lower` = Estimate - 1.96 * SE,
           `95% CI upper` = Estimate + 1.96 * SE)
}

out_ns  <- get_RD_RR(fit_ns)  %>% mutate(Model = "Primary (no‑switch)")
out_itt <- get_RD_RR(fit_itt) %>% mutate(Model = "Sensitivity (ITT)")

results <- bind_rows(out_ns, out_itt) %>%
  select(Model, Time, Estimand, Estimate, `95% CI lower`, `95% CI upper`)

knitr::kable(results, digits = 3,
             caption = "90‑day risk difference and risk ratio for AKI: primary (no‑switch) vs sensitivity (ITT) estimands – one‑step TMLE via *concrete*")
```

---

\### Diagnostics & plots
Optional: plot risk curves or RD using `plot()` on the `ConcreteOut` object; see package vignette for customisation.

```{r optional-plots, eval=FALSE}
# Uncomment to display counterfactual risk curves
# plot(getOutput(fit_ns, Estimand = "Risk"), ask = FALSE)
# plot(getOutput(fit_ns, Estimand = "RD"),  ask = FALSE, NullLine = TRUE)
```

---

\### Interpretation

* **Primary (no‑switch)** reflects the counterfactual scenario where switching is prevented; censoring‑related selection is addressed internally by TMLE → unbiased biologic effect of continuous SOF exposure.
* **Sensitivity (ITT)** answers the pragmatic policy question: what happens when clinicians *start* SOF vs non‑SOF, allowing subsequent regimen changes.

Consistency between the two estimates supports robustness; divergence would flag treatment‑effect dilution (cross‑over) or residual confounding of switching.

---

*Generated with the `concrete` R package (v≥0.1.0).*
