---
title: "Dynamic & Stochastic Adherence‑Forgiveness with **lmtp**"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

> **Goal.** Walk through a *complete* causal‐inference pipeline for the question:
>
> > *Is Antiretroviral Regimen&nbsp;1 virologically superior to Regimens 2–n when we account for dynamic adherence patterns?  At what adherence threshold is it still “forgiving” ?*
>
> We (i) simulate realistic longitudinal data, (ii) define static, dynamic **and** stochastic adherence interventions, and (iii) estimate 1‑year risk & risk‑differences with {lmtp} (v≥1.1).

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, echo = TRUE)

# BONUS: progressr progress bars!
#progressr::handlers(global = TRUE)
```

## 1 Road‑map for a Longitudinal Causal Question

| Road‑map Step | Our specification |
|--------------:|-------------------|
| **1a Question** | *Adherence forgiveness*: “Had everyone initiated Regimen 1 (vs. observed comparators) **and** we dynamically boosted each 90‑day block’s adherence (`PDCᵗ`) to ≥ ϑ, what is the 12‑mo risk of virologic failure?” |
| **Population** | Adults starting *any* first‑line ART in an integrated EHR/claims DB |
| **Treatment nodes `Aₜ`** | Indicator that current regimen is **Regimen 1** (1) or another comparator (0) at start of each 90‑day block (`t = 0,…,K−1`) |
| **Time‑varying `Lₜ`** | Last‑block % days‑covered (`PDCᵗ₋¹`), log viral‑load, creatinine, comorbidity flare score |
| **Outcome `Y`** | Viral‑suppression failure at 12 months (VL > 200 copies/ml) |
| **Censoring `Cₜ`** | Loss to FU, death, regimen discontinuation |
| **Estimand** | $\,\psi(\vartheta)=\Pr\bigl\{Y^{d_{\vartheta}}=1\bigr\}$ under the MTP that raises `PDCᵗ` to ≥ ϑ whenever below |
| **Identification** | Sequential exchangeability, positivity, consistency |
| **Estimation** | `lmtp_sdr()` (sequentially doubly‑robust) with **sl3** Super Learner |
| **Sensitivity** | Vary ϑ ∈ {0.5, 0.7, 0.9}; IPCW for LTFU; alternative VL thresholds |


## 2 Simulating a Realistic ART Cohort

We create 5 blocks (`K = 5`, i.e. 15 months) so the last viral‑load occurs ≈ 12 mo.

```{r simulate}
library(tidyverse)
library(data.table)

sim_dat <- function(N = 4000, K = 5, censor_prob = .15, seed = 123) {
  
  set.seed(seed)
  
  ## -------- baseline ---------------------------------------------------
  id  <- 1:N
  age <- rnorm(N, 45, 10)
  sex <- rbinom(N, 1, 0.4)
  cd4 <- pmax(50, rnorm(N, 500, 150))
  
  DT  <- CJ(id, t = 0:(K - 1))
  DT  <- merge(DT, data.table(id, age, sex, cd4), by = "id")
  
  ## -------- block 0 ----------------------------------------------------
  DT[t == 0, `:=`(
    A    = rbinom(.N, 1, plogis(0.5 - 0.01 * age + 0.5 * sex)),
    PDC  = runif(.N, .5, 1)
  )]
  
  ## -------- iterate blocks 1 … K-1 ------------------------------------
  for (tt in 1:(K - 1)) {
    
    ## lag from previous block
    lag <- DT[t == tt - 1,
              .(id,
                A_prev   = A,
                PDC_prev = PDC)]
    
    DT <- merge(DT, lag, by = "id", all.x = TRUE)
    
    ## treatment switching uses lagged vars
    DT[t == tt,
       A := rbinom(.N, 1,
                   plogis(1.5 * A_prev -
                            3 * (PDC_prev < .6) -
                            0.01 * age))]
    
    ## new current adherence
    DT[t == tt,
       PDC := pmin(1,
                   pmax(0,
                        rnorm(.N,
                              0.85 * A + 0.6 * (1 - A), .2)))]
    
    ## drop the temporary lags
    DT[, c("A_prev", "PDC_prev") := NULL]
  }
  
  ## -------- viral load & outcome --------------------------------------
  DT[, VL := rnorm(.N, 4 - 1.2 * A - 2 * PDC, 0.6)]
  
  Y <- DT[t == K - 1,
          .(id,
            Y = rbinom(.N, 1,
                       plogis(-3 - 0.4 * A - 2 * PDC + 0.3 * (VL > 4))))]
  
  ## -------- censoring --------------------------------------------------
  drop <- data.table(id,
                     t_cens = rbinom(N, 1, censor_prob) *
                       sample(1:K, N, TRUE))
  
  DT <- merge(DT, drop, by = "id")
  DT[, C := as.integer(t >= t_cens & t_cens > 0)]
  DT[, t_cens := NULL]
  
  ## -------- merge outcome, order, return ------------------------------
  long <- merge(DT, Y, by = "id")
  setorder(long, id, t)
  
  long <- as.data.frame(long)
  return(long[])
}

## test
long_dat <- sim_dat() %>% as.data.frame()

```


## 3 Define Shift Functions (Static, Dynamic, Stochastic)

```{r shifts}
library(lmtp)
# 3.1 Dynamic “boost-to-threshold” (MTP)
make_dyn <- function(thr=.8){
  function(data, trt){
    pdc <- data[[gsub("^A","PDC_prev", trt)]]
    at  <- data[[trt]]
    ifelse(pdc < thr, 1, at)   # force Regimen 1 when adherence low
  }
}

# 3.2 Stochastic shift: when pdc<thr, switch to Regimen 1 with prob p
make_stoch <- function(thr=.8, p=.7){
  function(data, trt){
    pdc <- data[[gsub("^A","PDC_prev", trt)]]
    at  <- data[[trt]]
    need<- (pdc < thr)
    flip<- rbinom(length(at),1,p)
    ifelse(need & flip==1, 1, at)
  }
}
```


## 4 Super Learner Library via **sl3**

```{r sl3}
library(SuperLearner)
sl_lib <- c("SL.glm")
```


## 5 Fit SDR‑TMLE under Natural Course & Interventions

```{r fit, eval=F}


# Map time‑varying covariates into list-of-lists for lmtp
Lnodes <- list( c("PDC"))

# Natural course (reference)
fit_nat <- lmtp_sdr(
  data          = long_dat,
  id            = "id", 
  trt           = "A",                 # single treatment col in long data
  outcome       = "Y",                 # third argument
  baseline      = c("age", "sex", "cd4"),
  time_vary     = Lnodes,              # list of time-varying covars
  cens     = "C",                 # name of censoring column
  outcome_type  = "binomial",
  folds         = 5,
  learners_trt      = sl_lib,
  learners_outcome  = sl_lib
)

# Dynamic thr=0.8
fit_dyn <- lmtp_sdr(
  data          = long_dat,
  id            = "id",
  trt           = "A",          # <-- one column
  outcome       = "Y",
  baseline      = c("age","sex","cd4"),
  time_vary     = Lnodes,       # list of "PDC" or "PDC_prev"
  cens     = "C",          # <-- one column
  shift         = make_dyn(.8),
  outcome_type  = "binomial",
  folds         = 5,
  learners_trt      = sl_lib,
  learners_outcome  = sl_lib
)

# Stochastic (thr=0.8, p=0.7)
fit_sto <- lmtp_sdr(long_dat, "Y", trt_nodes,
                    baseline = c("age","sex","cd4"),
                    time_vary = Lnodes,
                    censoring = cen_nodes,
                    shift = make_stoch(.8,.7),
                    outcome_type = "binomial",
                    folds = 5,
                    learners_trt = sl_lib,
                    learners_outcome = sl_lib)
```


## 6 Results

```{r results, eval=F}
print(summary(fit_nat))
print(summary(fit_dyn))
print(lmtp_contrast(fit_dyn, fit_nat, type = "diff"))
print(summary(fit_sto))
```

*Interpretation*   `lmtp_contrast()` shows the **risk difference** if we had ensured ≥80 % adherence each block.  A negative RD ⇒ fewer failures (better suppression) under the dynamic boost.


## 7 Sensitivity: Varying the Threshold

```{r sens, eval=F}
ths <- c(.5,.7,.9)
cts <- lapply(ths, function(th){
  f <- lmtp_sdr(long_dat,"Y",trt_nodes,
                baseline=c("age","sex","cd4"), time_vary=Lnodes,
                censoring=cen_nodes, shift=make_dyn(th),
                outcome_type="binomial", folds=5,
                learners_trt=sl_lib, learners_outcome=sl_lib)
  data.frame(th=th, psi = f$estimates$theta, se=f$estimates$std.error)
})
knitr::kable(do.call(rbind, cts), digits=3)
```


## 8 Discussion

* **Dynamic vs. stochastic**: Dynamic MTPs enforce a deterministic rule; stochastic shifts can mimic partial program uptake.
* **Why SDR/TMLE?** Doubly‑robust + efficient in large models; compatible for ≥90‑day blocks; easy Super Learner integration.
* **Limitations**: Requires positivity (`PDC` boosted for *some* low‑adherence patients in observed data). Outcome measured at a fixed time; extensions exist for restricted mean‑time‑to‑suppression.


## 9 References

1. Díaz, I., & van der Laan, M. (2021). *Non‑parametric causal effects based on longitudinal modified treatment policies.* JASA.
2. Young, J. G., et al. (2020). *A causal roadmap for generating high‑quality real‑world evidence.* JCTS.
3. Rudolph, K., et al. (2022). *lmtp: An R package for longitudinal modified treatment policies.*
