---
title: "HCV → AKI: 90‑Day Primary & Sensitivity Analyses"
output:
html_document:
toc: true
---

> **Purpose.** Estimate the 90‑day cumulative risk of first acute kidney injury (AKI) comparing (A = 1) sofosbuvir‑containing vs (A = 0) non‑sofosbuvir DAA regimens **under a hypothetical *no‑switch* intervention** (primary estimand, IPCW‑adjusted) and contrast it with two routinely reported sensitivities:
>
> \* **Treatment‑policy ITT** – ignores switching/discontinuation.
> \* **Naïve as‑treated** – censors 30 d after the first regimen switch (no weighting).

The simulated claims‑style dataset (\_sim\_hcv\_aki\_complex.csv) was generated with `DGP.R`.

---

\## 1  Setup

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(here)
library(survival)   # Kaplan–Meier & Cox
library(ipw)        # IPC weight creation
library(splines)    # flexible censoring model
```

\## 2  Load & tidy the data

```{r data-load}
df <- read_csv(here("data/sim_hcv_aki_complex.csv"), show_col_types = FALSE) %>%
  mutate(
    # key analysis variables
    time       = follow_time,          # already min(event, censor)
    event      = event,                # 1 = AKI, 0 = censored
    trt        = treatment,            # 1 = SOF, 0 = non‑SOF
    switched   = switch               # 1 if regimen switch occurred
  )

# restrict to ≤ 90 days for all estimands
cutoff <- 90
```

\## 3  Primary estimand – IPCW *no‑switch* (while‑on‑treatment)

\### 3.1 Define analytic censoring indicator
We artificially censor each subject at *min*(first switch day + 30, administrative 90 d). All post‑censor events are treated as missing via weighting.

```{r censor-def}
# administrative 90‑day horizon
admin_censor <- if_else(df$time > cutoff, 1, 0)
# switch‑based censoring (+30 d risk window simulated via DGP)
switch_censor <- if_else(df$switched == 1 & df$time > 0, 1, 0)

# overall censoring indicator for IPCW model
C <- pmax(admin_censor, switch_censor)   # 1 = censored before AKI
Tstar <- pmin(df$time, cutoff)           # observed/administrative time

df_ipcw <- df %>% mutate(C = C, Tstar = Tstar, event_ipcw = if_else(event==1 & time<=cutoff & C==0, 1, 0))
```

\### 3.2 Estimate censoring weights
A simple pooled‑logistic (discrete‑time) model conditional on baseline covariates suffices for illustration; more flexible spline‑based models are easily substituted.

```{r ipcw-est}
# baseline covariates used in DGP
covars <- c("age","ckd","cirrhosis","diabetes","bmi","sex_male")

# expand person‑day data up to censor/event/cutoff
expand_rows <- function(dat){
  purrr::map_dfr(seq_len(nrow(dat)), function(i){
    tibble(id = dat$id[i], trt = dat$trt[i],
           day = seq_len(dat$Tstar[i]),
           censored = if_else(day == dat$Tstar[i] & dat$C[i]==1,1,0),
           event    = if_else(day == dat$Tstar[i] & dat$event_ipcw[i]==1,1,0),
           !!!dat[i, covars])
  })
}

long <- expand_rows(df_ipcw)

# pooled‑logistic model for censoring.
fitC <- glm(censored ~ trt + ns(day,4) + ., family = binomial(), data = long)
long <- long %>%
  mutate(pC = predict(fitC, type = "response")) %>%
  group_by(id) %>%
  mutate(Surv_uncens = cumprod(1 - pC)) %>% ungroup()

weights <- long %>% group_by(id) %>% summarise(w = last(Surv_uncens))
df_ipcw <- df_ipcw %>% left_join(weights, by = "id")
# truncate extreme weights (optional balance/stability)
df_ipcw <- df_ipcw %>% mutate(w = pmin(w, quantile(w, .995, na.rm=T)))
```

\### 3.3 Weighted Kaplan–Meier & risk estimates

```{r km-ipcw}
km_ipcw <- survfit(Surv(Tstar, event_ipcw) ~ trt, data = df_ipcw, weights = w)

# extract 90‑day survival & compute risks
surv_90 <- summary(km_ipcw, times = cutoff)$surv
risk_90 <- 1 - surv_90
risk_tbl <- tibble(trt = c("non‑SOF","SOF"), risk = risk_90)

risk_tbl
risk_diff <- diff(rev(risk_90))   # SOF – non‑SOF
```

\## 4  Sensitivity 1 – Treatment‑policy ITT (ignore switch)

```{r km-itt}
km_itt <- survfit(Surv(pmin(time, cutoff), event) ~ trt, data = df)
surv_90_itt <- summary(km_itt, times = cutoff)$surv
risk_90_itt <- 1 - surv_90_itt
diff_itt <- diff(rev(risk_90_itt))
```

\## 5  Sensitivity 2 – Naive as‑treated (censor at switch, no weighting)

```{r km-naive}
# censor 30 d after switch as already in DGP (follow_time)
km_naive <- survfit(Surv(pmin(time, cutoff), event) ~ trt, data = df_ipcw[df_ipcw$C==0 | df_ipcw$event_ipcw==1,])

risk_90_naive <- 1 - summary(km_naive, times = cutoff, extend = TRUE)$surv
diff_naive <- diff(rev(risk_90_naive))
```

\## 6  Results summary

```{r results}
res <- tibble(
  Estimand = c("Primary: IPCW no‑switch","Sensitivity: ITT","Sensitivity: naïve as‑treated"),
  Risk_nonSOF = c(risk_90[1], risk_90_itt[1], risk_90_naive[1]),
  Risk_SOF    = c(risk_90[2], risk_90_itt[2], risk_90_naive[2]),
  Risk_Diff   = c(risk_diff, diff_itt, diff_naive)
)
knitr::kable(res, digits = 3, caption = "90‑day cumulative risk of first AKI")
```

**Interpretation.** The IPCW‑adjusted *no‑switch* analysis isolates the biologic effect of continuous SOF exposure, whereas ITT represents the real‑world initiation policy, and the naïve censor‑at‑switch gives an upper‑bound that may be biased if switching is informative.

