---
title: "Step 4 (Statistical Estimand) – Translating the HCV→AKI Identification into a Quantity We Can Estimate"
output:
html_document:
toc: true
toc_depth: 2
---

##  Statistical Estimand step

* **Identification (Step 3)** converts the *causal* estimand – a
  parameter of the unobservable counterfactual distribution – into
  *some* functional of the *observable* data **under a set of
  assumptions**.  The output of Step 3 is often called an
  **observed‑data functional**.
* **Statistical Estimand (Step 4)** chooses **one particular summary of
  that functional** (e.g.\ risk difference, risk ratio, RMST‑difference)
  and pins down *how* we will evaluate it (single time‑point vs.\ curve;
  marginal vs.\ conditional; additive vs.\ multiplicative scale).
  – This choice must also respect the audience’s needs (regulatory vs.
  mechanistic) and the estimator’s large‑sample properties.

> Think of Step 2 as *mapping the road* and Step 4 as *deciding exactly
> where to park the car*.

The Roadmap paper (Figure 1 on page 3) shows Step 2 “Observed Data →
Identifiability” followed by Step 4 “Statistical Estimand & Estimator”
citeturn6file0.

---

## 2  Observed‑data functionals inherited from Step 2

We carry forward two functionals derived in the Identification
section:

$$
\begin{aligned}
\Psi_{\text{NS}}(P) &= E\Big[ W_i(90)\,Y_i \Big] \\
\Psi_{\text{ITT}}(P) &= E\Big[\,E\{Y \mid A=a\}\Big]\_{a=1}^{a=0}
\end{aligned}
$$

* \$W\_i(90)\$ is the inverse‑probability‑of‑censoring weight that
  up‑weights uncensored individuals out to 90 days (switching, death,
  and disenrolment treated as censoring).
* \$Y\_i\$ indicates first AKI within 90 days.

---

## 3  Statistical estimands for this protocol

| Label                                                  | Notation                                                                          | Natural scale  | Description                                                                                               |
| ------------------------------------------------------ | --------------------------------------------------------------------------------- | -------------- | --------------------------------------------------------------------------------------------------------- |
| **Primary** – Hypothetical *no‑switch* risk difference | \$\mathrm{RD}*{\text{NS}} = \Psi*{\text{NS}}^{a=1} - \Psi\_{\text{NS}}^{a=0}\$    | additive       | 90‑day difference in weighted cumulative incidence if **all** patients were kept on their initial regimen |
| Sensitivity – Treatment‑policy (ITT) risk difference   | \$\mathrm{RD}*{\text{ITT}} = \Psi*{\text{ITT}}^{a=1} - \Psi\_{\text{ITT}}^{a=0}\$ | additive       | 90‑day difference in cumulative incidence when switching is allowed                                       |
| (Optional) Risk ratio on each scale                    | \$\mathrm{RR}= \Psi^{a=1}/\Psi^{a=0}\$                                            | multiplicative | Reported for regulatory completeness                                                                      |

Both estimands are marginal **population‑level** contrasts, suited to
regulatory safety questions.

---

## 4  Illustration with `concrete`

Below is a *template* R chunk (not executed here) demonstrating how the
primary estimand is passed to `concrete`.  `formatArguments()` encodes
“treat everybody with SOF” vs.\ “treat everybody with non‑SOF” as two
static interventions; censoring weights are handled automatically when
`EventType = "status"` includes 0 = censor.

```{r eval=FALSE}
ConcreteArgs_NS <- formatArguments(
  DataTable    = df_ns,               # censor at switch dataset
  EventTime    = "time",
  EventType    = "status",           # 1 = AKI, 0 = censored
  Treatment    = "trt",              # 1 = SOF, 0 = non‑SOF
  Intervention = 0:1,                # treat‑all strategies
  TargetTime   = 90,                 # single horizon
  TargetEvent  = 1,                  # AKI only
  CVArg        = list(V = 10)
)
ConcreteEst_NS <- doConcrete(ConcreteArgs_NS)
RD_NS <- getOutput(ConcreteEst_NS, Estimand = "RD")
```

A parallel chunk using `df_itt` (full follow‑up, ignoring switch) yields
`RD_ITT`.

---

## 5  Communicating the estimands

For a pharmaco‑epi audience unfamiliar with the Roadmap we recommend
reporting *both* estimands side‑by‑side, emphasising:

* **Policy relevance** – ITT mirrors “start‑regimen” decisions.
* **Biologic relevance** – No‑switch isolates on‑treatment toxicity but
  relies on IPC weighting and time‑varying positivity.

A short boiler‑plate sentence suitable for manuscripts:

> “The primary statistical estimand was the 90‑day risk difference in
> first acute kidney injury had all patients remained on their index
> antiviral regimen, estimated by continuous‑time one‑step TMLE using
> inverse‑probability‑of‑censoring weights.  A treatment‑policy (ITT)
> estimand served as a prespecified sensitivity analysis.”

---

## 6  Key take‑aways for regulators & clinicians

* Declaring the *statistical* estimand forces us to pick the exact
  contrast (RD vs RR, 90 d vs 120 d, etc.).
* The observed‑data functional is the *blueprint*; the statistical
  estimand is the *specific measurement* we will read off that
  blueprint.
* Both must be specified **before** looking at results.

---



# Old 

## Aim of Step 4
Translate each causal estimand (Step 1a) into a precise mathematical
statistical estimand—a functional of the observed data distribution
that equals the causal quantity if the identification assumptions
hold (Step 3).

* **Statistical estimand**  
  \[
    \Psi_{\text{RD}}(P_0)=
      \mathbb{E}_{W}\!\bigl[\, Q(90,1,W) - Q(90,0,W) \bigr],
  \]
  where  
  \(Q(t,a,W)=\Pr\{Y(t)=1 \mid A=a, W, C(t)=0\}\).

* **Planned estimator**  
  Discrete-time TMLE via `survtmle(method = "mean")` with SuperLearner libraries for both \(g\) (treatment & censoring) and \(Q\) (failure).



# 4.1 Notation recap

* \(O = (W,\,A,\,C,\,Y)\) – baseline covariates \(W\); treatment \(A \in \{0,1\}\); censoring indicator \(C(t)\); outcome \(Y(t)\) (AKI observed by day \(t\)).  
* Distribution of \(O\): \(P_0\).  
* Statistical expectations with respect to \(P_0\): \(\mathbb{E}_{P_0}[\cdot]\).

We operate on the discrete-time grid \(t = 1,\dots,90\) (days).

# 4.2 Linking causal ↔ statistical estimands


| Label | Causal estimand (potential-outcome form) | Statistical estimand (functional of \(P_0\)) |
|-------|------------------------------------------|----------------------------------------------|
| **ΔRisk\(_{0–90}\)**<br>(treatment-policy) | \(\psi_{\text{RD}} = \mathbb{E}\!\bigl[Y^{(1)}\!\le 90\bigr] - \mathbb{E}\!\bigl[Y^{(0)}\!\le 90\bigr]\) | \(\displaystyle \Psi_{\text{RD}}(P_0)=\mathbb{E}_{W}\!\left[\, Q(90,1,W) - Q(90,0,W) \right]\) |
| **Cox HR**<br>(while-on-treatment) | Instantaneous hazard ratio while on initial regimen | \(\displaystyle \Psi_{\text{HR}}(P_0)=\exp\{\beta_A\}\) where \(\beta_A\) solves the Cox partial-likelihood score |
| **ΔRMST(0–90)** | \(\psi_{\text{RMST}} = \int_{0}^{90}\! \bigl[S^{(1)}(u)-S^{(0)}(u)\bigr]\,du\) | \(\displaystyle \Psi_{\text{RMST}}(P_0)=\sum_{u=1}^{89}\! \bigl[S(u,1)-S(u,0)\bigr]\) |
| **Hypothetical no-switch ΔRisk** | Same as ΔRisk but under intervention “never switch” | \(\displaystyle \Psi_{\text{NS}}(P_0)=\mathbb{E}_{W}\!\bigl[ Q^{\dagger}(90,1,W) - Q^{\dagger}(90,0,W) \bigr]\) |
| **Principal-stratum ΔRisk** | ΔRisk among subjects who would adhere ≥ 8 wk under either arm | \(\displaystyle \Psi_{\text{PS}}(P_0)=\frac{\mathbb{E}\!\bigl[ \pi(W)\,[Q(90,1,W)-Q(90,0,W)]\bigr]}{\mathbb{E}[\pi(W)]}\) with \(\pi(W)=\Pr(A^{\!*}=1,\,A^{\#}=0\mid W)\) |

*Definitions used in the functionals*

\[
\begin{aligned}
Q(t,a,W)        &= \Pr\bigl\{Y(t)=1 \mid A=a,\,W,\,C(t)=0\bigr\},\\
S(t,a)          &= \Pr\bigl\{Y(t)=0,\,C(t)=0 \mid A=a\bigr\},\\
Q^{\dagger}(t,a,W) &= \text{Risk under the “never-switch” intervention},\\
\pi(W)          &= \Pr\bigl\{\text{adhere} \mid W\bigr\}.
\end{aligned}
\]

Each statistical estimand is a functional of the observed-data
distribution \(P_0\); under the identification assumptions
(Step 3) it equals its corresponding causal estimand.

# 4.3 Example: computing the statistical estimand in the simulation

### 4.3.1 Plug-in (g-formula) estimate

```{r}
library(tidyverse)
dat <- read.csv(here::here("data/sim_hcv_aki.csv"))

risk <- dat |>
  group_by(A = treatment) |>
  summarise(risk = mean(event == 1 & follow_time <= 90))

plug_RD <- diff(rev(risk$risk))   # risk_A=1 – risk_A=0
plug_RD



```

# 4.4 Mapping estimands to planned estimators (preview of Step 5)


| Statistical estimand | Planned primary estimator | Double-robust? | Uses machine learning? |
|----------------------|---------------------------|---------------|------------------------|
| **ΔRisk\(_{0–90}\)** (treatment-policy) | `survtmle` (`method = "mean"`) | **Yes** | **Yes** (SuperLearner for \(g\) and \(Q\)) |
| **Cox HR** (while-on-treatment) | PS-matched Cox with robust SE | **No** | **No** (legacy baseline model) |
| **ΔRMST(0–90)** | Aalen-Stratified stacked-survival TMLE | **Yes** | **Yes** |
| **ΔRisk (hypothetical no-switch)** | (i) Censor-at-switch TMLE<br>(ii) Switch-hazard TMLE | **Yes** | **Yes** |
| **Principal-stratum ΔRisk** | Substitution estimator with sensitivity band | **No** | Optional (HAL or GLM) |

# 4.5 Stakeholder checks




## Take-home
Step 4 anchors each causal question to a concrete, data-level
functional. Every subsequent choice of estimator, diagnostic and
sensitivity analysis flows from these definitions.
