---
title:  "Roadmap Step 1b — Causal Model"
date:   "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    code_folding: hide
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, fig.align = "center")
```

# Study-type & background knowledge

Step 1b in the causal roadmap explicitly defines the statistical model and discusses assumptions necessary for causal identification. Clear articulation of these assumptions is crucial for valid causal inference.


## Target-trial emulation schematic
We emulate a two-arm pragmatic trial that randomises at t = 0 (first DAA dispense) to

A = 1 Sofosbuvir-containing regimen

A = 0 Non-sofosbuvir regimen

and follows individuals for 90 days to the first occurrence of acute kidney injury (AKI) or censoring.
The analytic data used here are fully simulated by generate_hcv_data() in DGP.R, with parameters calibrated to the real HealthVerity HCV cohort. No protected health information is present.

*Roadmap link:* this schematic translates Step 1a’s causal question into an explicit target-trial design 

## Causal graph (baseline DAG)


```{r}

library(tidyverse) 
library(dagitty) 
library(ggdag)

# DAG specification -------------------------------------------------------
#   W  = baseline confounders
#   A0 = initial DAA regimen (SOF vs non-SOF)
#   C  = censoring processes (death, switch, disenroll)
#   Y  = 90-day AKI

dag_txt <- "
dag {
  W  -> A0
  W  -> Y
  A0 -> Y
  W  -> C
  A0 -> C
}"

aki_dag <- dagitty(dag_txt)

# Visualise ---------------------------------------------------------------
ggdag(aki_dag, text = FALSE, use_labels = 'name', seed = 123) +
  theme_dag() +
  ggtitle('AKI Safety DAG: Sofosbuvir vs Non-Sofosbuvir')
```


**Explanation of DAG components:** 

- **W (Baseline Covariates):** Factors like age, sex, baseline kidney function, diabetes, liver cirrhosis, healthcare utilization. These influence both treatment selection (A) and AKI outcome (Y). 
- **A (Treatment):** Initial choice between SOF-containing or non-SOF-containing DAAs. 
- **Y (Outcome):** AKI event within 90 days after treatment initiation.
- **C (Censoring events):** Events that lead to incomplete follow-up, such as death, regimen switching, or loss of insurance. The DAG explicitly assumes no direct arrows from unknown or unmeasured variables to A and Y (conditional on W), reflecting the key assumption of conditional exchangeability. 


The DAG encodes our current subject-matter understanding: baseline factors influence both treatment choice and AKI risk; treatment may induce early kidney events and may also trigger regimen switching or dropout (informative censoring).




## Intercurrent events & time-varying mechanisms

| Intercurrent event | Representation | Roadmap implication |
|--------------------|---------------|---------------------|
| **Regimen switch** (non-SOF → SOF or vice-versa) | First crossover time enters `C(t)` | Competes with AKI; requires clear *estimand* choice (treatment-policy vs hypothetical no-switch). |
| **Death** | Cause-specific hazard in `C(t)` | Competing risk; can be handled via composite outcome or Fine–Gray estimand. |
| **Loss of follow-up / disenrollment** | Administrative censoring | Treated as random given measured `W`; verify positivity and apply IPC weighting if needed. |

Because the current dataset is *simulated* with baseline treatment assignment only (no programmed switching), time-varying confounding is absent **by design**. This simplification lets us focus on demonstrating the Roadmap mechanics without additional longitudinal complexity. Note the additional simulated case study at the end of this tutorial demonstrating longitudinal TMLE to untangle longitudinal confounding in HIV treatment adherence.


\## 1.2 Specifying the Causal Model (DAG) Causal inference typically involves representing relationships between variables visually using directed acyclic graphs (DAGs). A DAG helps clarify the assumptions necessary for valid causal inference, particularly: - **Exchangeability (no unmeasured confounding)** - **Positivity (treatment assignment possible across covariates)** - **Consistency (treatment definitions correspond to reality)** Consider the following simplified DAG illustrating relationships in our AKI study:

```{r dag, echo=FALSE, eval=FALSE}


dag_txt <- " dag { W -> A W -> Y W -> C A -> Y A -> C }" 
ggdag(dagitty(dag_txt), text=FALSE, use_labels='name', seed=873) + theme_dag() + ggtitle('DAG: Sofosbuvir (SOF) vs Non-SOF DAAs and AKI risk')
```

## Identification conditions  

1. **Exchangeability:** \(Y^{a} \perp\!\!\!\perp A \mid W\).  
2. **Positivity:** \(0 < P(A = a \mid W = w) < 1\) for all \(w\) in support \(W\).  
3. **Consistency:** Observed \(Y\) equals \(Y^{a}\) for the treatment actually received.  
4. **Correct model for censoring:** \(C\) independent of counterfactual outcomes conditional on \(W, A\).

These minimal assumptions will be revisited formally in Step 3 (Identifiability Assessment).




### Link to simulation design

| Feature in the simulated DGP | Rationale for inclusion | Estimands that depend on it |
|------------------------------|-------------------------|----------------------------|
| **Baseline covariates \(W\) generated under exchangeability** | Guarantees the causal assumptions for ΔRisk and ΔRMST hold by design, allowing validation of estimator bias. | All estimands |
| **No programmed regimen switching** | Ensures ITT and “no-switch” estimands coincide in simulation so any divergence in real data highlights switching bias. | Diagnostic reference for ITT, hypothetical no-switch, while-on-Tx |
| **Event-time generator calibrated to 2 % 90-day AKI risk** | Provides realistic incidence for variance demonstrations and power calculations. | All estimands (affects standard errors and coverage) |

#### Note on fitness-for-purpose audit
A “fitness-for-purpose” data audit is normally completed at this stage to verify that variable definitions, time-stamps, and measurement reliability align with the causal model. Because we are using fully simulated data whose generating mechanism is known and transparent, such an audit is not applicable. When we later port the workflow to the real HealthVerity cohort, this audit will be mandatory.






**Outcome-Blind Sandbox Approach**: Conducting all modeling choices (covariate selection, learner tuning, etc.) prior to outcome unblinding or using permuted/simulated datasets ensures unbiased model selection, avoiding overfitting or biased results.







