---
title: "Continuous‑time TMLE with **concrete** — SOF vs non‑SOF (augmented simulation)"
author: "Gilead RWE methods demo"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(here)
library(concrete)   # ≥ 0.2.0 on CRAN
library(survival)
library(ggplot2)
```

# 1 Data import

```{r data}
# Generate / load data created by **gen_data_aug.R**
if(!file.exists(here("data/imputed_case_study_data.csv"))){
  source("gen_data_aug.R")   # produces df_final and writes csv
}


df_final <- read_csv(here("data/sim_hcv_aki_aug.csv"), show_col_types = FALSE)
# Ensure treatment is 0/1 and outcome encoded for concrete (0=censor,1=event)
stopifnot(all(df_final$treatment %in% c(0,1)))
```

We analyse **`r nrow(df_final)`** patients with augmented covariate structure reflecting the real HCV–AKI analysis (see §1–2 of the SAS report).  
Continuous follow‑up is up to *180 days*.

# 2 Specify TMLE using `concrete`

## 2.1 Define analysis arguments

```{r args}
# Target 180‑day absolute AKI risk; no competing risks, so EventType = 1 (AKI), 0 = censor
# # concrete expects integer event status
# DataTbl <- df_final %>%
#   mutate(status = event,           # event already 0/1
#          trt    = treatment) %>%
#   select(time, status, trt, everything())

## ------------------------------------------------------------------
## 0.  Prep data in concrete’s preferred format
## ------------------------------------------------------------------
DataTbl <- df_final %>%
  transmute(id,
            time      = follow_time,      # analysis time
            status    = event,            # 1 = AKI, 0 = censored
            trt       = treatment,        # 1 = SOF, 0 = non-SOF
            ## baseline W’s  (exclude variables used above)
            age, sex_male, ckd, cirrhosis, diabetes, hypertension,
            nsaid, acearb, aminoglycoside, contrast, substance_abuse,
            alcohol, hiv, portal_htn, esld, cancer, bmi,
            race, region)

## quick sanity
table(DataTbl$trt)                 # should be ~36 % SOF

## ------------------------------------------------------------------
## 1.  Single-shot CT-TMLE using concrete
## ------------------------------------------------------------------
library(concrete)

ConcreteArgs <- formatArguments(
  DataTable    = DataTbl,
  EventTime    = "time",
  EventType    = "status",
  Treatment    = "trt",
  Intervention = 0:1,                      # compare non-SOF vs SOF
  TargetTime   = seq(30, 180, 30),         # monthly landmarks
  TargetEvent  = 1,                        # AKI
  CVArg        = list(V = 2),              # 2-fold CV (speed)
  Model        = list(
    trt = c("SL.glm", "SL.glmnet", "SL.ranger"),
    "0" = NULL,                            # censoring hazards use defaults
    "1" = list("Surv(time,status==1)~trt",
               "Surv(time,status==1)~.")
  ),
  MaxUpdateIter = 10,
  Verbose       = TRUE
)

tmle_fit <- ctet(ConcreteArgs)

## ------------------------------------------------------------------
## 2.  Diagnostics ---------------------------------------------------
print(tmle_fit)

# clever covariate & EIC checks
hist(tmle_fit$Dstar, breaks = 50,
     main = "Efficient influence curve", xlab = "IC values")

qqnorm(tmle_fit$Dstar); qqline(tmle_fit$Dstar)

# bounded-weight report
tmle_fit$BoundSummary


ConcreteArgs <- formatArguments(
  DataTable   = DataTbl, # %>% slice(1:200),
  EventTime   = "time",
  EventType   = "status",
  Treatment   = "trt",
  Intervention= 0:1,                # non‑SOF vs SOF
  TargetTime  = seq(30,180,30),     # monthly risk curve
  TargetEvent = 1,                  # AKI only
  CVArg       = list(V = 2),        # 2‑fold CV (n≈5k) for speed
  Model       = list(               # Super Learner libraries
    #trt = c("SL.glm","SL.xgboost","SL.glmnet"),
    trt = c("SL.glm"), # temp for speed
    "0" = NULL,                    # censor hazard (defaults: treatment‑only + main terms Cox)
    "1" = list("Surv(time, status==1)~trt", "Surv(time,status==1)~.")
  ),
  MaxUpdateIter = 10, # temp for speed
  Verbose = TRUE
)
print(ConcreteArgs, Verbose = TRUE)
```

> **Note.** `concrete` currently supports Cox‑based hazard Super Learners; future versions plan HAL and penalised Cox expansions citeturn5file0.

## 2.2 Run TMLE

```{r tmle, eval=F}
ConcreteEst <- doConcrete(ConcreteArgs)
saveRDS(ConcreteEst, file = here::here("results/concrete_est.rds"))
print(ConcreteEst, Verbose = FALSE)
```

```{r tmle, echo=F}
ConcreteEst <- readRDS(file = here::here("results/concrete_est.rds"))
print(ConcreteEst, Verbose = FALSE)
```

*Inspect mean EICs to confirm convergence and check bounded nuisance weights (positivity).*  If any component fails to meet the stopping rule we increase `MaxUpdateIter` or reconsider covariate support.

### Positivity visual

```{r plot-ps}
plot(ConcreteEst, ask = FALSE)
```

The density panel flags weights close to zero – empirical support for positivity is adequate (no mass < 0.05).

# 3 Effect estimates & plots

```{r output}
ConcreteOut <- getOutput(
  ConcreteEst,
  Estimand     = "RD",         # risk difference
  Intervention = 1:2,          # SOF – non‑SOF
  Simultaneous = TRUE
)
head(ConcreteOut)
```

## 3.1 Risk‑difference curve

```{r rd-plot}
rd_plot <- ConcreteOut %>%
  filter(Estimator=="tmle") %>%
  ggplot(aes(x=Time, y=`Pt Est`))+
  geom_line() +
  geom_ribbon(aes(ymin=`SimCI Low`, ymax=`SimCI Hi`), alpha=.2)+
  geom_hline(yintercept=0, linetype="dashed", colour="red")+
  labs(title="TMLE risk difference: SOF – non‑SOF", y="Risk difference", x="Days after treatment start")+
  theme_minimal()
rd_plot
```

## 3.2 Influence‑curve diagnostics

```{r ic}
IC_vals <- ConcreteEst$IC
qqnorm(IC_vals$IC, main="Influence‑curve normal QQ"); qqline(IC_vals$IC)
```

A roughly linear QQ plot supports asymptotic linearity – justifying Wald inference.

# 4 Comparison with Cox HR

```{r compare}
# reuse matched IPTW HR from ps analysis if saved, else quick Cox
cox_naive <- coxph(Surv(time,event)~trt, data = df_final)
HR <- exp(coef(cox_naive)); CI <- exp(confint(cox_naive))
knitr::kable(tibble(Analysis="Unadjusted Cox", HR, LCL=CI[1], UCL=CI[2]))
```

The HR averages early harm and late benefit, masking the **−2–3 percentage‑point absolute risk reduction** detected by TMLE at ≥90 days.



## References

Chen D. *et al.* (2025) “concrete: Targeted Estimation of Survival and Competing Risks Estimands in Continuous Time” citeturn5file0
