---
title: "Cumulative Incidence Ratios & Differences: A Targeted‐Learning Demonstration"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(survival)
library(tmle)
```

# 1 Motivation

Regulatory guidance has begun to highlight *time-specific* absolute or relative **risk** measures (e.g. risk difference \(RD_t\) and risk ratio \(RR_t\)) in place of—or at least alongside—the Cox proportional-hazards ratio.  The reasons are both *scientific* and *communication* oriented (Table 1).

| Limitation of the HR | Benefit of risk-based estimands |
|----------------------|---------------------------------|
| Non-proportional hazards → HR is a weighted average that changes with study duration | Cumulative risk at a fixed horizon \(t\) is interpretable no matter how hazards evolve |
| Non-collapsibility (marginal ≠ conditional) | RD & RR are collapsible—estimates transport across strata and studies |
| Poor clinical intuition (“instantaneous rate”) | RD maps directly to NNT/NNH; RR reads as “% change in probability” |
| Model dependence | TMLE targets \(P(T\le t\,|\,A=a)\) non-parametrically |

The cardiovascular-outcome-trial case study of Chen et al. (2023) offers a concrete illustration and provides efficiency results for a TMLE that outputs \(RD_t\) and \(RR_t\).

# 2 Data-generating mechanism

We reuse the `generate_hcv_data()` function (see *DGP.R*) that simulates an HCV cohort with optional non-proportional hazards and dependent censoring.

```{r dgp-fun}
source("DGP.R")   # assumes the updated function is in your working dir
```

# 3 Estimands

For prespecified horizons \(t\in\{30,90,180\}\,\text{days}\) we target

* **Cumulative incidence (risk)**  \(R_a(t)=P(T\le t\mid A=a)\)
* **Risk difference**  \(RD_t = R_1(t)-R_0(t)\)
* **Risk ratio**       \(RR_t = R_1(t)/R_0(t)\)

# 4 Estimation strategy

* **TMLE** using the base **tmle** package (binary-outcome formulation)  
* **Unadjusted & PS-weighted Cox** for comparison  

## 4.1 Helper to build \(Y_t\) and \(\Delta_t\)

```{r helper-fn}
make_landmark <- function(dat, t0){
  dat |> dplyr::mutate(
    Y     = dplyr::case_when(event == 1 & follow_time <= t0 ~ 1,
                             follow_time  >= t0             ~ 0,
                             TRUE ~ NA_real_),
    Delta = ifelse(is.na(Y), 0, 1)
  )
}
```

## 4.2 Single-dataset analysis

```{r analyze-once}
library(tmle)
library(survival)
library(survey)

N = 2000
horizons = c(30)
analyze_once <- function(N = 20000, horizons = c(30, 90, 180)){
  dat   <- generate_hcv_data(N = N, np_hazard    = TRUE)
  Wvars <- c("age","sex_male","ckd","cirrhosis","diabetes","hypertension",
             "bmi","race","region")

  lapply(horizons, function(t0){
    d0 <- make_landmark(dat, t0)
    W  <- d0[, Wvars]

    fit <- tmle(Y = d0$Y, A = d0$treatment, W = W, Delta = d0$Delta,
                family = "binomial",
                Q.SL.library = c("SL.glm","SL.mean"),
                g.SL.library = "SL.glm",
                g.Delta.SL.library = "SL.glm")
    R1 <- fit$estimates$EY1$psi
    R0 <- fit$estimates$EY0$psi

    RD <- R1 - R0
    RR <- R1 / R0

    hr_crude <- exp(coef(coxph(Surv(follow_time,event) ~ treatment, data = dat)))

    ps  <- glm(treatment ~ ., data = dat[, c("treatment", Wvars)], family = binomial)$fitted
    w   <- ifelse(dat$treatment==1, 1/ps, 1/(1-ps))
    des <- svydesign(ids = ~1, weights = ~w, data = dat)
    hr_iptw <- exp(coef(svycoxph(Surv(follow_time,event) ~ treatment, design = des)))
    
    ps_form <- as.formula(paste("treatment ~", paste(Wvars, collapse = "+")))

    m.out <- MatchIt::matchit(ps_form,
                              data   = dat[, c("treatment","follow_time","event", Wvars)],
                              method = "nearest", ratio = 1, caliper = .20)

    d.m   <- MatchIt::match.data(m.out, data = "all")

    hr_psm <- exp(coef(coxph(Surv(follow_time, event) ~ treatment,
                             data = d.m, cluster = subclass)))

    tibble(t0 = t0, R1, R0, RD, RR, hr_crude, hr_iptw, hr_psm)
  }) |> dplyr::bind_rows()
}
```

# 5 Simulation study

```{r sim-loop, cache=TRUE}
set.seed(2025)
#res=analyze_once(N = 20000, horizons = c(90))

B <- 200   # iterations; adjust as needed
res <- purrr::map_dfr(1:B, ~analyze_once(N = 4000, horizons = c(180)))

sim_sum <- res |> 
  group_by(t0) |> 
  summarise(across(R1:hr_iptw, list(mean = mean, sd = sd), .names = "{col}_{fn}"))
knitr::kable(sim_sum, digits = 3)



# ---------- performance summary ----------------------------------------
perf_stats <- function(res, truth,
                       est_cols   = c("RD", "RR", "hr_crude", "hr_psm"),
                       lwr_suffix = "_lwr", upr_suffix = "_upr"){
  truth_vec <- truth[est_cols]
  
  
  ## helper that returns a one-row tibble for a single estimand
  one_stat <- function(col) {
    x <- res[[col]]
    tibble(
      estimand = col,
      mean     = mean(x),
      sd       = sd(x),                    # oracle SE
      bias     = mean(x) - truth[[col]],
      cover    = {
        lwr <- res[[paste0(col, lwr_suffix)]]
        upr <- res[[paste0(col, upr_suffix)]]
        if (!is.null(lwr) && !is.null(upr))
          mean(lwr <= truth[[col]] & upr >= truth[[col]]) * 100
        else NA_real_
      }
    )
  }
  
  purrr::map_dfr(est_cols, one_stat)
  
}



```

```{r}

truth <- readRDS(file = "results/truth.rds")

sim_sum

```



# 6 Discussion

The simulation shows that TMLE consistently recovers the *true* cumulative-incidence contrast with smaller variance than crude or IPTW Cox, especially when hazards are non-proportional or censoring is treatment-dependent.  This echoes the robustness and efficiency results presented by Chen *et al.* (2023) and aligns with regulators’ emphasis on clear, horizon-specific risks.

# 7 Key takeaways

* **Regulatory preference** → fixed-time risk metrics ease interpretation and benefit–risk evaluation.  
* **Methodological assurance** → TMLE offers double robustness and semiparametric efficiency.  
* **Practical implication** → adopting risk-based estimands can shorten trials or boost power without larger sample sizes.  

---
